Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-25 paper/section-discussion.tex
intro, contributions

Geometric theorem
proving is another %possible 
application. Harrison
\cite{harrison09without} comments on the pervasiveness of % usefulness of
invariance properties in this area. Programs in our framework
automatically satisfy invariance properties, removing the need for
{\em ad hoc} proofs. % of these facts.

@@

\paragraph{Semantic Equality} In general, the semantic equality
$\stackrel{sem}\sim$ in \defref{def:semantic-equality} is not an
equivalence relation. If the interpretations of all 
primitive types are partial equivalence relations then semantic
equality is indeed an equivalence relation. However, this excludes the
geometry and distance-indexed examples. More generally, we can
consider relational interpretations that are \emph{difunctional}. (A
relation is difunctional if whenever $(x,y)$, $(x',y')$ and $(x,y')$
are in the relation then so is $(x',y)$.) Difunctionality is weaker
than being a PER, but still suffices to prove that semantic equality
is an equivalence relation. Hofmann \cite{hofmann08correctness} has
used difunctional relations in the setting of effect
analyses. Difunctionality covers all our examples except
distance-indexed types. Note that for both PERs and difunctional
relations we need to close the relational interpretation of
existential types under the appropriate property to ensure that all
types are interpreted as PERs/difunctional relations. For
distance-indexed types it is possible that a new notion of
equivalence based on closeness is required.

>>>>

Geometric theorem
proving is another %possible 
application. Harrison
\cite{harrison09without} comments on the pervasiveness of % usefulness of
invariance properties in this area. Programs in our framework
automatically satisfy invariance properties, removing the need for
{\em ad hoc} proofs of these facts.

@@

\paragraph{Semantic Equality} In general, the semantic equality
in \defref{def:semantic-equality} is not an
equivalence relation. If the interpretations of all 
primitive types are partial equivalence relations then semantic
equality is indeed an equivalence relation. However, this excludes the
geometry and distance-indexed examples. More generally, we can
consider relational interpretations that are \emph{difunctional}. (A
relation is difunctional if whenever $(x,y)$, $(x',y')$ and $(x,y')$
are in the relation then so is $(x',y)$.) Difunctionality is weaker
than being a PER, but still suffices to prove that semantic equality
is an equivalence relation. Hofmann \cite{hofmann08correctness} has
used difunctional relations in the setting of effect
analyses. Difunctionality covers all our examples except
distance-indexed types. Note that for both PERs and difunctional
relations we need to close the relational interpretation of
existential types under the appropriate property to ensure that all
types are interpreted as PERs/difunctional relations. For
distance-indexed types it is possible that a new notion of
equivalence based on closeness is required.

### 2012-10-25 paper/section-introduction.tex
intro, contributions

Relational parametricity is in essence a principle of
\emph{invariance}: the behaviour of polymorphic code is invariant
under changes of data representation. Invariance results also abound in
mathematics and physics. The area of a triangle is invariant with
respect to isometries of the Euclidean plane; the determinant of a
matrix is invariant under changes of basis; and Newton's laws are the
same in all inertial frames. Typically, transformations in mathematics
and physics have interesting structure; for example, translations in
the Euclidean plane form an abelian group.

%Types express properties of programs. The statement $\mathit{b}:
%\tyBool$ tells us that $\mathit{b}$ has value $\mathtt{true}$ or
%$\mathtt{false}$; the statement
%$\mathit{pos}:\tyReal\tyProduct\tyReal$ says that $\mathit{pos}$ is a
%pair whose components are real numbers. More interestingly, the
%statement $\mathit{f} : \forall
%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that if we apply
%$\mathit{f}$ to a list whose elements satisfy some property $\psi$,
%then the result is a list whose elements also satisfy $\psi$.

%Unary predicates only go so far, and for polymorphic types
%particularly do not express everything that follows from possessing
%that type.  Imagine a function $\mathit{oddRev}$ that reverses any
%list whose first element is an odd number, otherwise behaving as the
%identity. It certainly preserves properties of the elements -- but
%truly parametric polymorphism should tell us more.

%To obtain stronger properties we must \emph{relate} multiple
%executions of the same code. The statement $\mathit{f}:\forall
%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
%representation of the list elements supplied to $\mathit{f}$ is
%reflected in the elements in the result.  Formally, for any function
%$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
%\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
%essence of \emph{relational parametricity} as formulated by
%Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
%immediately tells us that $\mathit{oddRev}$ cannot be given a
%polymorphic type: simply consider the change-of-representation given
%by $\lambda x. x+1$.  But more profoundly it can be used to prove
%blah blah (initial algebras, term reps, non-inhabitation, etc.).

In previous work the third author studied relational parametricity for
types parameterized by units-of-measure, whose invariance 
properties relate to changes of units, or \emph{scaling}~\cite{kennedy97relational}. 
In this paper we investigate the general case of types indexed by attributes
with algebraic structure. For example, in computational geometry, points in
the plane can be indexed by a coordinate frame; in information-flow
security, computations are indexed by principals; in differential privacy, types
can be indexed by `distance'. Types that are
polymorphic over such indices induce invariance properties and
abstraction barriers beyond those introduced by their unindexed
versions, as we now illustrate.

\paragraph{Invariance}
A function $\mathit{areaTri}$ that computes the area of a triangle
can be assigned the type:
$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
But we can also assign it the following more expressive polymorphic
type:
\[
\mathrm{areaTri} : \forall t\mathord:\SynTransl{2}.
  \tyPrim{vec}{t} \times \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyReal \\
\]
This type expresses the fact that if each of the arguments to $\mathrm{areaTri}$
is translated by the same vector, then the result remains the same,
that is, it is \emph{invariant} under translation. Formally, for any 
vector $\vec t$,
\[
\mathrm{areaTri}\;(\vec{t} + \vec{v_1}, \vec{t} + \vec{v_2}, \vec{t} + \vec{v_3}) = 
\mathrm{areaTri}\;(\vec{v_1}, \vec{v_2}, \vec{v_3})
\]

Transformations typically \emph{compose} in various
ways, and the compositions satisfy algebraic laws. For example, 
we can assign a function that computes the area of a circle given its
radius the following polymorphic type:
\[
\mathrm{areaCircle} : \forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to
\tyPrim{real}{s\cdot s}
\]
This captures the fact that the area of a circle varies as the square
of its radius, i.e., $\mathrm{areaCircle}(k r) = k^2\cdot
\mathrm{areaCircle}(r)$ for any $k\neq 0$ (i.e., any $k$ in the
one-dimensional general linear group $\GL{1}$).  Here, $s$ 
can be interpreted as the \emph{units of measure} of the argument to
$\mathrm{areaCircle}$, and `$\cdot$' composes units using the
product. We can also add an inverse operation and identity unit of
measure $1$, and then impose the algebraic laws of abelian
groups. This permits identification of, for example,
$\tyPrim{real}{s\cdot s^{-1}}$ with the type
$\tyPrim{real}{1}$ of dimensionless constants.

@@

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall t\mathord:\SynTransl{2}.\tyPrim{vec} t\to
\tyPrim{vec} t$ are related when they are related at type
$\tyPrim{vec} t\to\tyPrim{vec}t$ for all translations
$\vec t\in\Transl{2}$ associated with $t$.

\vspace*{0.1in}

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. As is the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall s\mathord:\SynGL{1}.\tyPrim{real} s\to\tyPrim{real} s)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s\cdot s}\to\tyPrim{real}s$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}s$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(s)$ for a positive real number
$\rho(s)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as %a type:
%\begin{displaymath}
 $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' abstraction theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}

>>>>

Relational parametricity is in essence a principle of
\emph{invariance}: the behaviour of polymorphic code is invariant
under changes of data representation. For example, the type
$\forall\alpha.\tyPrim{list}\alpha\to\tyPrim{list}\alpha$
tells us that any transformation applied to elements of
the input list will be reflected by the same transformation applied
to elements of the result. 
Invariance results also abound in
mathematics and physics. The area of a triangle is invariant with
respect to isometries of the Euclidean plane; the determinant of a
matrix is invariant under changes of basis; and Newton's laws are the
same in all inertial frames. Typically, the 
transformations under which invariants are preserved have interesting structure: 
for example, translations in
the Euclidean plane form an abelian group.

Inspired by this connection, we study type systems that
capture rich invariants through types indexed by attributes with
algebraic structure.  For example, in computational geometry, points
in the plane can be indexed by attributes representing affine
transformations; in information-flow security, computations can be
indexed by principals; in differential privacy, types can be indexed
by `distance'. Types that are polymorphic over such indices induce
invariance properties and abstraction barriers beyond those introduced
by their unindexed versions, as we shall illustrate.  This generalises
previous work by the third author on types parameterized by
units-of-measure, whose invariance properties relate to changes of
units, or \emph{scaling}~\cite{kennedy97relational}.

Our motivations for studying algebraically-indexed types are
threefold. First, we believe that, as with
units-of-measure~\cite{fsharp}, practical programming language
extensions will follow. For example, in computational geometry and
graphics, attributes on points, vectors, and other geometric types
could be used to prevent the mixing of different coordinate systems,
or `frames'. Second, type-based static analyses can be based on
indexed types, for example, in effect systems~\cite{benton06reading},
and, more speculatively, in continuity
analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
expressing algebraic invariants through types has the potential to
offer slick proof techniques for mechanized
mathematics. Harrison has applied the
invariance properties of geometric primitives to create elegant proofs
in geometry, based on `without loss of
generality' principles~\cite{harrison09without}. The invariance properties are expressed 
and propagated using ad-hoc tactics; our types offer a more principled
means of achieving the same end, and the `wlog' principle itself
is expressed through type isomorphisms.

We follow the mantra \emph{semantics first, syntax later} in studying
types with algebraic structure. We have not yet built a practical
programming language that supports algebraically-indexed types; nor
have we designed type checking, type inference, or static analysis
algorithms.  But when we do so, the semantics will guide us. The fact
that zero is polymorphic in units-of-measure (it can be given type $\forall
u.\tyPrim{real}u$) whereas other constants are dimensionless (having
type $\tyPrim{real}1$) is justified by the invariance properties
induced by the types: zero is invariant under scaling, other constants
are not. For less trivial constants and operations, the appropriate types are
not so apparent, as we shall see, but invariance properties expressed by
the semantics guide us in assigning appropriate types. Semantics does not lie!

\paragraph{Invariance}
To illustrate type-induced invariance properties, consider
2-dimensional geometry. 
A function $\mathit{areaTri}$ that computes the area of a triangle
can be assigned the type:
$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
But we can also assign it the following more expressive polymorphic
type:
\[
\mathrm{areaTri} : \forall t\mathord:\SynTransl{2}.
  \tyPrim{vec}{t} \times \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyReal \\
\]
This type expresses the fact that if each of the arguments to $\mathrm{areaTri}$
is translated by the same vector, then the result remains the same,
that is, it is \emph{invariant} under translation. Formally, for any 
vector $\vec t$,
\[
\mathrm{areaTri}\;(\vec{t} + \vec{v_1}, \vec{t} + \vec{v_2}, \vec{t} + \vec{v_3}) = 
\mathrm{areaTri}\;(\vec{v_1}, \vec{v_2}, \vec{v_3})
\]

Transformations typically \emph{compose} in various
ways, and the compositions satisfy algebraic laws. For example, 
we can assign a function that computes the area of a circle given its
radius the following polymorphic type:
\[
\mathrm{areaCircle} : \forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to
\tyPrim{real}{s\cdot s}
\]
This captures the fact that the area of a circle varies as the square
of its radius, i.e., $\mathrm{areaCircle}(k r) = k^2\cdot
\mathrm{areaCircle}(r)$ for any $k\neq 0$ (the `sorts' $\SynTransl{2}$ and
$\SynGL{1}$ will be explained later).  Here, $s$ 
can be interpreted as the \emph{units of measure} of the argument to
$\mathrm{areaCircle}$, and `$\cdot$' composes units using the
product. We can also add an inverse operation and identity unit of
measure $1$, and then impose the algebraic laws of abelian
groups. This permits identification of, for example,
$\tyPrim{real}{s\cdot s^{-1}}$ with the type
$\tyPrim{real}{1}$ of dimensionless constants.

@@

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}s$ are related when $v$ ``scales to'' $w$
according to an interpretation of $s$ (i.e., $w=\rho(s)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall t\mathord:\SynTransl{2}.\tyPrim{vec} t\to
\tyPrim{vec} t$ are related when they are related at type
$\tyPrim{vec} t\to\tyPrim{vec}t$ for all translations
$\vec t\in\Transl{2}$ associated with $t$.

As it happens, the relational interpretations given above are
functional, relating one value uniquely to another. Other applications
make use of primitive relations that are not simple functions.
For example, in a type system in which the index in
$\tyPrim{real}s$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(s)$ for a positive real number
$\rho(s)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as %a type:
%\begin{displaymath}
 $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' abstraction theorem, for a given \emph{model}
of index sorts and relational interpretation of primitive types.
We prove that the semantics soundly approximates contextual equivalence.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\item
We improve on the earlier semantics of
units-of-measure~\cite{kennedy97relational} in a number of ways.  By
extending the language of units with an `absolute value' operation, we
can give more precise types and obtain more general invariance
properties.  The relational interpretation for units is both simpler
and more flexible, and we derive slicker proofs of non-definability,
and new results.  Our notion of type isomorphism is stronger than
previously, being based on contextual equivalence.
\end{itemize}
We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}
